Data Management \\ and LSST Special Programs

### 2017-09-01 ms.tex
major revision

\author{M.~L.~Graham and M.~Juri\'{c}}

@@

\setDocAbstract{\textbf{WORKING DRAFT (contains rough notes and partial thoughts) } but intended to be an overview of any and all potential requirements on data management (DM) from Special Programs such as the deep drilling fields (DDF) or mini-surveys. We also summarize the plans for how the Special Programs data will be integrated into the wide-fast-deep (WFD) Main Survey (Levels 1 and 2). It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future Special Program white papers and preparations for Level 3 pipelines. }

@@

\noindent The purpose of this study is to: \\
(1) summarize the current plans of the DM pipelines with respect to Special Programs, \\
(2) summarize the processing that is required to enable science for Special Programs, \\
(3) assess whether the DM pipelines meet the expected needs of Special Programs, and \\
(4) ensure any necessary changes get written into the requirements, designs, and plans.

\noindent This will be accomplished by: \\
(1) starting with the DPDD, whitepapers, etc., and compiling relevant information, \\
(2) discussing initial issues internally with DM (Mario, the SST), engineering (K-T, Tim), and then \\
(3) ensuring the identified needs become specified requirements (i.e., spawn change requests).

{\it Current Status: This document is a rough draft, with the goal of being exhaustively complete. Many of the identified \textbf{``Concerns"} may just be misunderstandings or a lack of knowledge.}

The rest of this introduction contains: \\
\S~\ref{ssec:intro_bounds} -- defining DM-related boundaries for Special Programs. \\
\S~\ref{ssec:intro_forum} -- existing channels for interaction with Science communities. \\
\S~\ref{ssec:intro_UC} -- a summary of unresolved \textbf{Concerns} that arose in this study.

Later sections attempt to identify potential \textbf{Concerns} regarding DM's plans for processing the data from Special Programs with a variety of approaches: \\
\S~\ref{sec:dmdocs} -- a review of the current DM plans for SP data that are either already explicitly written down in DM documentation, or exist in Mario's thoughts. \\
\S~\ref{sec:issues} -- a brainstorm of potential issues that might arise, organized by pipeline or product, inspired by internal and external documentation. \\
\S~\ref{sec:science} -- a review of science cases for SP and their potential strain on the DM system, including detailed reduction/analysis case studies in \S~\ref{ssec:science_dmsums}.

\clearpage
% % % % % % % % % % % % % % % % % % 
\subsection{Boundaries of Allowed Special Programs Observations}\label{ssec:intro_bounds}

In the next couple of years, cadences and pointings for DDF and new mini-surveys will be proposed by the community. The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018. We will need to define boundaries for these observations from a DM prospective. The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3}.

Start of a list of DM-related considerations for Special Programs data. \\
\textbf{Exposure times --} (e.g., current shortest is 1 s, with stretch goal of 0.1, but... \\
\textbf{Afternoon calibrations --} (e.g., must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained) \\
\textbf{Time allotted --} Nominally, 10\% of the total observing time (90\% goes to the WFD).

% % % % % % % % % % % % % % % % % % 
\subsection{Interacting with the Science Community} \label{ssec:intro_forum}

To resolve some open issues we will need to solicit input from the science collaborations (SC; also, remember that the SC Coordinator for DDF is Neil Brandt). It may be better and easier to wait until the next call for community input, and request that a "DM Requirements" section be included in all white papers (as in \S~\ref{ssec:science_dmsums}). Also, there is already a community forum on DDF, so far minimally used but ready to be revived when the next calls soliciting white papers are released. \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.

@@

\clearpage
% % % % % % % % % % % % % % % % % % 
\subsection{A Summary of Unresolved Concerns} \label{ssec:intro_UC}

\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C1} Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C2} Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C3} Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C4} Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C5} Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications. \end{enumerate}
\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C6} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C11} Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C12} Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the data restrictions for ISR. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C13} Will there be a Science Platform capability set up so that users are notified not just when any data for their Special Program is taken, but when certain sets are finished and ready for processing? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C14} Does the Level 3 pipeline have to include the steps to regenerate the processed exposures, or will DM deliver a service to automatically initiate regeneration by e.g., simply calling for an exposure or difference image? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C15} Is it yet known how users will apply for batch processing priority and how they will be notified their jobs have finished, e.g., RSS feed? I could imagine that this is a requirement that would be defined in a future document on the Science Platform deliverables? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C17} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C18} Confirm that the idea of federated Level 3 codes still exists (e.g., given how the conversation on incorporating a photo-$z$ code ended up in a proposed descope). \end{enumerate}

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\clearpage
\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmdocs}

In this section we summarize how DM intends to process the data obtained during Special Programs observations. This high-level "intent" is a combination of specifications that are actually written down in DM-related documents, and unwritten plans that exist in e.g., Mario's Brain. {\it MLG: Note that Sections \ref{ssec:dmdocs_SPinWFD} and \ref{ssec:dmdocs_L3} currently contain a mixture of what is officially written down and what is unofficially planned or expected, usually my interpretation of a conversation with Mario, while Section \ref{ssec:dmdocs_review} is a direct review of the DM documentation. Once the SST and engineering have converged on a treatment plan for Special Programs data via this work, my plan is to return to the official DM documentation and identify where change requests need to be initiated.}

>>>>

\author{M.~L.~Graham, M.~Juri\'{c}, and E.~Bellm}

@@

\setDocAbstract{\textbf{WORKING DRAFT (contains rough notes and partial thoughts) } but intended to be an overview of any and all potential requirements on data management (DM) from Special Programs such as the deep drilling fields (DDF) or mini-surveys. We also summarize the plans for how the Special Programs data will be integrated into the wide-fast-deep (WFD) Main Survey data products(Levels 1 and 2). It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future Special Program white papers and preparations for Level 3 pipelines.}

@@

The intent of LSST DM for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):

\vspace{-20pt}
\begin{quote}
LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
\end{quote}

\noindent The purpose of this study is to ensure that the requirements and plans of DM, as written down in all major DM documentation, accurately reflects this intent. In particular, this study is designed to review: \\
(1) the processing and products that are required to enable science for Special Programs, \\
(2) the current plans for DM processing and products with respect to Special Programs, \\
(3) whether the DM plans meet the expected needs of Special Programs, and \\
(4) any necessary changes to be written into the requirements, designs, and plans.

\noindent The concerns raised by this study fall into three main categories: \\
(1) Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
(2) Ensuring that DM's intent to process Special Programs data with reconfigured pipelines and provide separate data products that meet the science goals of each program is accurately reflected in all documents, as it is in the DPDD. \\
(3) Clarifying the computational sizing plans for processing and storage of Special Programs data and whether it is this separate from the allocation for end-user Level 3 analysis.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\section{The Potential Diversity of Special Programs Data} \label{ssec:data}

To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data. The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (standard visit images), then it might not be adequately handled by the DM system. To get a handle on the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes (Section \ref{ssec:data_bounds}). Next, we explore the current set of proposed science programs for LSST (Section \ref{ssec:data_science}) and the types of observations they would want to do, and then describe when and how we can incorporate revisions to these proposals from the user community in the future (Section \ref{ssec:data_comm}). Finally, we explore whether DM can or should set down any technical constraints on how far an observing program can deviate from a sequence of standard visit images and still be processed by DM pipelines, be they the reconfigured and/or the standard Level 1 and 2 pipelines (Section \ref{ssec:data_DMconstraints}).

% % % % % % % % % % % % % % % % % % 
\subsection{Observing Boundaries for Special Programs}\label{ssec:data_bounds}

We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals. So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}. \cite{2008arXiv0805.2366I} describes a nominal DDF data set as e.g., a series of fifty 15 second exposures in each of the $griz$ filters obtained once every two nights for four months, which would yield a nightly stack depth of $r<26.5$ and a full stack depth of $r<28.0$ (once combined with the wide-fast-deep survey data). Additionally, three mini-survey areas that have already been considered in detail are likely to move forward: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \cite{2008arXiv0805.2366I}). We have a summary collection of past white papers and informally proposed mini-surveys in Section \ref{ssec:data_science}.

All other aspects of LSST Special Programs remain open to proposals, including: \\
$\bullet$ additional deep drilling fields \\
$\bullet$ refined observing strategies for deep drilling fields \\
$\bullet$ optimized survey areas for the NES, South Pole, and Galactic Plane \\
$\bullet$ refined observing strategies for the NES, South Pole, and Galactic Plane \\
$\bullet$ additional mini-surveys \\
$\bullet$ the total fraction of time spent on Special Programs (nominally 10\%)

There will be some technical limitations on the mini-surveys that will be possible. Here's a rough draft list of such potential issues. It is not the job of this document to define the camera/observatory technical boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing. This list will be updated in the future (or revised to point to where these boundaries are defined). In Section \ref{sssec:dmplans_review_oss}, as we review the Observatory System Specifications (OSS, LSE-30, \cite{LSE-30}) document we highlight a couple more questions regarding potential boundaries on Special Programs observations.

Filter Changes (and see \ref{OSS-6}): \\
$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293). \\
$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances. However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes. \\
$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights.

Exposure Times (and see \ref{OSS-5}): \\
$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
$\bullet$ The maximum exposure time is not restricted. \\
$\bullet$ However, for exposure times there are other considerations. The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question. Changing the exposure time also affects the photometric and astrometric calibrations. Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag. A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations. Additionally, the impact on CR rejection routines is untested for long exposures.

Camera Rotation (and see \ref{OSS-7}): \\
$\bullet$ \textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism? Can't remember where I heard it.}

Telescope Slews: \\
$\bullet$ \textit{MLG - Obviously, long slews would cause additional overheads and a loss of observing efficiency, but are there other limitations?}

Calibration Images (and see \ref{OSS-2}, \ref{OSS-4}): \\
$\bullet$ The necessary afternoon calibrations for a Special Program must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained.

Operations Simulations: \\
$\bullet$ Mini-survey observing strategies will have to show that they do not negatively impact the LSST's ability to complete the wide-fast-deep survey; not necessarily in the next call for white papers, but later during the review and revise stage, once the new OpSim runs incorporating proposed mini-surveys have been generated.

Total Time: \\
$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey), but this could be changed if significant science is enabled. \\

% % % % % % % % % % % % % % % % % % 
\subsection{Currently Proposed Special Programs} \label{ssec:data_science}

In this section we compile information about the science goals and observational methods for Special Programs that have already been proposed. We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science; this Section is mainly descriptive and we do not spawn any {\bf Concerns} here. In Section \ref{sec:SPCS}, we create much more detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs; any {\bf Concerns} that arise are detailed in that section. Below, we list the set of resources that we used to compile this information, and in Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified along with an {\it incomplete} list of potential mini-surveys that people are thinking about (mainly from the presentations of Brandt and Ridgway at the 2016 AHM).

\noindent {\bf Resources:} \\
$\bullet$ ``LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
$\bullet$ The LSST Science Requirements Document (SRD), LPM-17, \cite{LPM-17} \\
$\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
$\bullet$ ``General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
$\bullet$ ``Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
$\bullet$ ``LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
$\bullet$ Chapter 10 ``Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}

\textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al. (2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 ``Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months. Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.

\medskip
\noindent \textbf{Solar System Objects (SSO)}\\
$\bullet$ North Ecliptic Spur: observations in this area of the sky will yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (Brandt talk). \\
$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \cite{2017arXiv170804058L}) \\
$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) 
$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{2017arXiv170804058L}) \\
$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}. Regarding SAS, \cite{BeckerWP} says that ``the multi-fit algorithm ... naturally provides a base infrastructure for this process. In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required. However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." \\
$\bullet \bullet \bullet$ {\bf Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night. The exception is a Sweetspot survey, which occurs during twilight and thus may require shorter exposures. Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 {\tt DIASource} catalogs updated each night. The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3. See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_MM} for more detailed DM processing case studies.

\noindent \textbf{Stars in the Milky Way and Magellanic Clouds} \\
$\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations. When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}. \\
$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}. For this, special co-adds may be required, e.g., {\it "to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}. \citep{SzkodyWP}. \\
$\bullet$ A Twilight Short Exposure survey would enable bright stars, which have longer monitoring baselines due to historical observations, to be put on the same photometric system as the deeper LSST WFD main survey catalog. \\
$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}. \\
$\bullet \bullet \bullet$ {\bf Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars. Similarly, most of these science goals can be met by reconfiguring the planned DM pipelines to detect sources, measure proper motion, and create deep stacks. The exception might be in processing exposures as short as 0.1 seconds, and attempting to measure flux of a saturated stars. See Sections \ref{ssec:SPCS_GPVSEx} and \ref{ssec:SPCS_M67} for more detailed DM processing case studies.

\noindent \textbf{Exoplanets} \\ 
$\bullet$ Transits. The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits. A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have ${\rm SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \cite{2008arXiv0805.2366I}). \cite{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane.\\
$\bullet$ Microlensing.  Slower than a transit, \cite{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates, which would then need follow-up with external facilities. However, this will require image differencing in crowded fields. \cite{2013arXiv1304.3455G} suggests that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped. Microlensing can also be done with DDF-style observing at low Galactic latitudes. \\
$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images. They might be possible with reconfigured DM pipelines, but this depends heavily on performance in crowded fields. See Section \ref{ssec:SPCS_GPVSEx} for a more detailed DM processing case study for Galactic Plane regions.

\noindent \textbf{Supernovae} \\
$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses. The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$. \cite{KesslerWP} \\
$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \cite{2014ApJ...794...23D}. \\
$\bullet$ Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template. This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days. These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss. SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science. The latter, specifically the ``characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}. \\
$\bullet$ \textit{MLG - I'm not sure this has been yet proposed, but short-exposure observations to include nearby SNe on the same photometric system could be useful?} \\
$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe). Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data. In particular, the Level 2 DRP codes to create ``transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images. See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.

\noindent \textbf{Galaxies} \\
$\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects. \cite{FergusonWP} mentions ``identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and ``characterization of low-surface-brightness extended features around both nearby and distant galaxies". \\
$\bullet$ The DDF stacks can also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. \\
$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP,GawiserWP}. \\
$\bullet \bullet \bullet$ {\bf Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources. In addition, it seems likely that Level 3 algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, but these are beyond the scope of DM.

\noindent \textbf{Weak Lensing} \\
$\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
$\bullet$ {\it Jim Bosch -- ``Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
$\bullet$ {\it Jim Bosch -- ``Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline. We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} ({\tt \Large{Community}} forum)  \\
$\bullet$ {\it MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
$\bullet \bullet \bullet$ {\bf Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.

% % % % % % % % % % % % % % % % % % 
\subsection{Future Input from Data Rights Holders} \label{ssec:data_comm}

Community input on LSST Special Programs will be ingested through a formal call for white paper proposals (Section \ref{sssec:data_comm_call}). This call will request that these white papers include a section that describes the processing needs that will enable their science goals; examples of processing case studies for some proposed Special Programs are given in Section \ref{sec:SPCS}. The DM team will also interact informally with the community through, e.g., a {\tt \Large{Community}} forum (Section \ref{sssec:data_comm_forum}) and might learn about users needs for DM processing through that means.

\subsubsection{Call for White Paper Proposals}\label{sssec:data_comm_call}

The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018. At the 2017 All Hands Meeting the Science Advisory Council seemed likely to recommend these be combined into a single call. The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3} and the Project's initial thoughts regarding this call for white papers are described by Beth Willman at the LSST SAC Meeting in October 2016\footnote{\url{https://project.lsst.org/groups/sac/sites/lsst.org.groups.sac/files/Willman_DDF.pdf}}. The plan for ingesting these white paper proposals is for them to be reviewed by the Science Advisory Council based on a set of criteria set by the Project Science Team, and the SAC will make recommendations to the LSST Director. The criteria to be met by the proposals may include specifications that they: satisfy minimum technical requirements (feasibility, overheads); maximize diverse scientific objectives (serve a wide community); and generate legacy datasets and add value to the products of other facilities. After the proposals have been reviewed (and perhaps culled or combined, as recommended by the SAC) it is projected that new OpSim runs will be produced in mid-2019 for further evaluation. While the initial white paper proposals will be separate from the existing Observing Strategy White Paper \cite{2017arXiv170804058L}, they could be incorporated into a comprehensive analysis of the options when these simulations are ready.

\textbf{Initial community suggestions regarding the call for white papers.} \\
$\bullet$ The call should describe the reconsideration timescale for evaluating Special Programs' progress and the power of the Project/SAC to terminate/initiate a Special Program. \\
$\bullet$ The call should be clear about whether a single white paper can be focused on a single science goal or a single observing strategy or, for example, to what extent they should incorporate all possible science for a given proposed set of observations. Sample outlines for white papers that are driven by a single science goal, or by a novel survey strategy, might help the proposers. \\
$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an ``eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
$\bullet$ Proposers will be asked to include "Data Management Considerations for Special Programs" into their white papers, and so they will also need a readable description of DM's plans and boundaries for Special Programs data (i.e., what this document is attempting to build).

\subsubsection{{\tt \Large{Community}} Forum}\label{sssec:data_comm_forum}

As the science community prepares their white paper proposals they have been invited to interact with Data Management through this {\tt \Large{Community}} forum on DDF. I am a watcher on this forum. So far it has been minimally used. \\ \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.

@@

% % % % % % % % % % % % % % % % % % 
\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:data_DMconstraints}

It might be helpful for the community to know, at the time of the call for new white papers, if certain non-standard visit images cannot processed by the DM pipelines. Exposures which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD. We consider shorter and longer exposure in turn.

(1) Shorter Exposures. The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint. The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint. Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale). As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for ``well-formed"). Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.

\begin{figure}
\begin{center}
\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin. At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center. We can see that for exposure times greater than 2 seconds, the centroid converges near its true value. \label{fig:expt}}
\end{center}
\end{figure}
